MLH1 Results Outline

### 2020-02-26 doc/Results.rmd
cleaning up Q1.1CO and Q1.IFD

We focus on the foci positions from one crossover (1CO n= 
) bivalents since the landscape patterns for multi crossover bivalents will be highly influences by crossover interference.

We anchor the position measures at the centromere (and normalize by deviding the position by the length of the chromosome, so the value range from 0 to 1, with smaller values representing positions closer to the centromere end of the bivalent, and larger values meaning foci positions closer to the opposite end (refered here as terminal or telomeric position).

We plot the mouse averages across all of the single bivalent measures, with the strain averages shown as boxplots...

in order to get a better picture of the sex differences in this foci-position metric

@@

In the domesticus strains the mouse averages for normalized single foci positions were significantly sexually dimorphic (fitting the predictions/pattern of the 'typical recombination landscape' (Sardell)). The single CO positions are medially placed in females and terminall placed in males.

The pavlues across the strains are; (t.test; p = `r Q1_1CO.pos_t.test.WSB$p.value`, `r Q1_1CO.pos_t.test.G$p.value`, and `r Q1_1CO.pos_t.test.LEW$p.value` for WSB, G and LEW respectively).

@@

After confirming the significan sex differences for 1CO position, we apply the familiar models

@@

<we don't explore the effects on the redcued bivalent data sets! The main expected pattern is confirmed>

Chromosome size effects are a confounding factor for CO position, with the expectation that shorter chromosomes will have a more uniform landscape compared to larger/longer chromosomes (cite). To account for this we again use the reduced bivalent sets (long.biv and short.biv)

@@

These box plot show that females have a much more medial position of single foci bivalents, (much closer to 50% compared to males). They also show that Musc males' Foci1 position is slightly more central / medial compared to the same type of positions in the Dom male strains. MOLF males have much more medial positions than other strains.

@@

```{r show.siscoten, echo=FALSE}
#think of a good way to display this

@@

```{r show.norm.siscoten, echo=FALSE}
Curated_BivData$nrm.siscoten <- Curated_BivData$SisCoTen/Curated_BivData$chromosomeLength

@@

```{r log.reg_siscoten, echo=FALSE}

@@

```{r telo.dist, echo=FALSE}

@@

```{r MM.telo, echo=FALSE}
#HetC

@@

```{r log.reg_telo.dist, echo=FALSE}

@@

```{r t.test_Telo.Dist, echo=FALSE}

@@

```{r cent.dist_show, echo=FALSE}
cent.dist_boxplots <- ggplot(Curated_BivData, aes(x= strain, y=dis.cent ) )+ geom_boxplot(aes( fill=hand.foci.count, group=strain), alpha=.2) + facet_wrap(sex~subsp, ncol=2, scales="free")+ggtitle("raw centromere distance")+ ylim(c(0, 150))

@@

```{r log.reg_CENT.dist, echo=FALSE}

@@

```{r t.test_CENT.Dist, echo=FALSE}
#CENT.DIST
#mouse av
t.test_mouse.av_Cent.dist <- t.test(

@@

```{r Q1.1CO.pos_Sardell.eq, echo=FALSE, include=FALSE}
#IGNORE THIS FOR KNOW, THIS

@@

```{r Q1.mean.nrmF1_table, echo=FALSE,eval=FALSE}

@@

In order to test if there are sex differences between crossover interference, a major determinant of the positioning of crossovers along chromosomes, we examined ... ran models with the interfocal distance (IFD) of two foci on the same bivalent.

-We focus on observations from two-crossover bivalents for more comparable observations.
-comparisons of all IFDs of mulit crossover bivalents is also included
Interference is a major determinant of the positioning of chromosomes

@@

DeBoer 2006, 13 -- measured chrm 1 in males and females, both were 2.8 microns
<!--
ToDo: make a melt.DF with the IFD1, IFD2, IFD3 in a single cols

explain the difference between the raw and PER measures. Why should I use /examine the IFD.PER?
-->

@@

The raw IFD measures -- are usuful for thinking of crossover interference as a mechnical - force mediated through the SC -- In some models this metric should be thought of as the best metric for charaecterizing crossover interference -- bc it reflect the physcial --

The above plots display the raw and normalized area which seperate two foci on the same chromosome. In the normalized plot, 9 observations fall above the length threshold of 150.  The female measures fall more uniformly across the range of (0 to 1).

@@

@@

-- run these comparisons to see if the small NRM IFD are due to shorter chromosomes, which are predicted to have weaker interference.

Looked at the KAZ pattern

@@

Mixed model analysis for IFD (interference), the first set of models are made with the lme() functions.

@@

The table above should display the slightly unusual pattern, where the coefficients for the significant sex fixed effect are positive and negative in the raw and normalized values respectively. That is for the raw IFD values, females are significantly longer but the normalized IFD values, males are significantly longer.

I tested 2 versions of the mixed model for this flavor of trait, raw IFD and normalized IFD measure. The tables below are from anova( for the lmer model ). Random effect of strain is not significant for ABS IFD, and only slightly significant for the IFD.PER

@@

the interaction effects were slightly significant for both raw and nrm.IFD.

The most significant value was the sex effect for nrm.IFDs.

The the random strain effect was not significant for either model.

For the raw measures, in M2, I think the 2 SKIVE effects mean that the female raw IFD is shorter than the male IFD.raw. The other effects are for PWD and SKIVE (larger raw IFD from intercept.)  I think the MSM and MOLF interaction effects were too far down the list to sop up any variance. For the M3, only SKIVE*male effect is close to significant

@@

Dive deeper into the sex specific pattern for each strain.  Below are code chunks which show the unusual sex specific results for IFD measures. The general pattern is that, female raw IFD > male IFD and  female PER IFD < male PER IFD. The scatter plots show that female raw measures are longer than male and for the PER values, the female mean is brought down by an enrichment of short IFDs.

For some strains, PWD, MSM and SKIVE there's a 30% threshold in the male PER IFD distributions. (What does that mean?). How do I test / quantify this pattern? Cluster metric?

@@

Above is a table of the proportion of 2Co bivalents which have a norm IFD below 30%, For all strains but KAZ, the females have a greater proportion of these shorted IFD values.

>>>>

We focus on the foci positions from one crossover (1CO n= `r `) bivalents since the landscape patterns for multi crossover bivalents (2COs) will be highly influenced by crossover interference.

The CO positions are normalized by the SC length and anchored at the centromere with the normalized position values ranging from 0 to 1, (refered here as terminal or telomeric position).

@@

In the domesticus strains the mean normalized foci positions were significantly different between sexes and fit the predictions of the 'typical recombination landscape' (with female positions being medially placed and terminally placed in males). The p.values across the strains are; (t.test; p = `r Q1_1CO.pos_t.test.WSB$p.value`, `r Q1_1CO.pos_t.test.G$p.value`, and `r Q1_1CO.pos_t.test.LEW$p.value` for WSB, G and LEW respectively).

@@

After confirming the significan sex differences for 1CO position, we apply the model frameworks

@@

REMOVE THE SISCOTEN PLOTs...

<ONLY SHOW 1 PLOT, we don't explore the effects on the redcued bivalent data sets! The main expected pattern is confirmed>

<!-- Ignore this section / kind of results for know
Chromosome size effects are a confounding factor for CO position, with the expectation that shorter chromosomes will have a more uniform landscape compared to larger/longer chromosomes (cite). To account for this we again use the reduced bivalent sets (long.biv and short.biv) -->

@@

@@

```{r show.siscoten, echo=FALSE, eval=FALSE}
#think of a good way to display this

@@

```{r show.norm.siscoten, echo=FALSE, eval=FALSE}
Curated_BivData$nrm.siscoten <- Curated_BivData$SisCoTen/Curated_BivData$chromosomeLength

@@

```{r log.reg_siscoten, echo=FALSE, eval=FALSE}

@@

```{r telo.dist, echo=FALSE, eval=FALSE}

@@

```{r MM.telo, echo=FALSE, eval=FALSE}
#HetC

@@

```{r log.reg_telo.dist, echo=FALSE, eval=FALSE}

@@

```{r t.test_Telo.Dist, echo=FALSE, eval=FALSE}

@@

```{r cent.dist_show, echo=FALSE, eval=FALSE}
cent.dist_boxplots <- ggplot(Curated_BivData, aes(x= strain, y=dis.cent ) )+ geom_boxplot(aes( fill=hand.foci.count, group=strain), alpha=.2) + facet_wrap(sex~subsp, ncol=2, scales="free")+ggtitle("raw centromere distance")+ ylim(c(0, 150))

@@

```{r log.reg_CENT.dist, echo=FALSE, eval=FALSE}

@@

```{r t.test_CENT.Dist, echo=FALSE, eval=FALSE}
#CENT.DIST
#mouse av
t.test_mouse.av_Cent.dist <- t.test(

@@

```{r Q1.1CO.pos_Sardell.eq, echo=FALSE, eval=FALSE, include=FALSE}
#IGNORE THIS FOR KNOW, THIS

@@

```{r Q1.mean.nrmF1_table, echo=FALSE, eval=FALSE}

@@

@@

DeBoer 2006, 13 -- measured chrm 1 in males and females, both were 2.8 microns -->

@@

Crossover interference is one of the major determinates -- of the CO landscape. We use X number of 2CO bivalents, to test if there are diffferences between males and females in interference. 
The mean interfocal distance (IFD) for raw and normalized values was calculated from X female and X male mice.

The raw IFD measures -- are usuful for thinking of crossover interference as a mechnical force mediated through the SC. Previous measures found no sex difference in IFD in B6 mice (de Boer et al )

@@

<remove the chrm size effects -- or check out model with chromosome length>

@@

@@

Since the most significant t.test were significantly different across sexes for the NORMALIZED effects -- we follow up with the model framework to --check out the potential evolutionary patterns--

We applied the same frame work of models, Mixed model analysis for IFD (interference), the first set of models are made with the lme() functions.

@@

@@

 (IFD.PER)
sex
MM p = `r`

subsp
p = `r`

the interaction effects were slightly significant for both raw and nrm.IFD.
The most significant value was the sex effect for nrm.IFDs. 
The the random strain effect was not significant for either model.

<For the raw measures, in M2, I think the 2 SKIVE effects mean that the female raw IFD is shorter than the male IFD.raw. The other effects are for PWD and SKIVE (larger raw IFD from intercept.)  I think the MSM and MOLF interaction effects were too far down the list to sop up any variance. For the M3, only SKIVE*male effect is close to significant>

@@

@@

<!-- junk for Q1.IFD

The above plots display the raw and normalized area which seperate two foci on the same chromosome. In the normalized plot, 9 observations fall above the length threshold of 150.  The female measures fall more uniformly across the range of (0 to 1).
-->
